---
title: "Free Throw Rate and Tournament Odds"
description: |
  Do teams that make it to the NCAA tournament get to the free throw line more often than those that don’t? Using the cbb.csv dataset, which includes Division I men’s basketball stats, I examine whether a team’s free throw rate (FTR) — calculated as free throw attempts per field goal attempt — is linked to postseason qualification. A permutation test is used to simulate what we’d expect if free throw rate and tournament selection were unrelated.
author: Henok Dawite
date: Apr 2, 2025
format: html
execute:
  warning: false
  message: false
toc: true
number-sections: true
---

## Introduction

This project looks into whether NCAA Division I men’s basketball teams that qualify for the tournament tend to reach the free throw line more often than those that don’t. We use Free Throw Rate (FTR), calculated as free throw attempts divided by field goal attempts (FTA/FGA), as a proxy for offensive aggressiveness. Teams with higher FTRs are likely getting fouled more and playing physically, and the question is whether that shows up in who makes the tournament.

## Data Cleaning and Preparation

```{r}
library(tidyverse)
library(janitor)
library(purrr)

cbb <- read_csv("cbb.csv") |> 
  clean_names() |> 
  mutate(made_tourney = if_else(postseason %in% c("2ND", "Champions", "E8", "F4", "R32", "R64", "R68", "S16"), "yes", "no")) |> 
  select(ftr, made_tourney) |> 
  filter(!is.na(ftr))
```

## Visualizing the Differences

```{r}
# Create a boxplot to compare FTR between tournament and non-tournament teams
ggplot(cbb, aes(x = made_tourney, y = ftr, fill = made_tourney)) +
  geom_boxplot() +
  labs(
    title = "Free Throw Rate by Tournament Status",
    x = "Made NCAA Tournament",
    y = "Free Throw Rate (FTA/FGA)")
```

## Observed Difference

```{r}
# Calculate observed difference in mean FTR between the two groups
obs_diff <- cbb |> 
  group_by(made_tourney) |> 
  summarise(mean_ftr = mean(ftr)) |> 
  summarise(diff = mean_ftr[which(made_tourney == "yes")] - mean_ftr[which(made_tourney == "no")]) |> 
  pull(diff)

obs_diff
```

**Interpretation:**\
The observed difference in mean Free Throw Rate between tournament and non-tournament teams is approximately 0.0135 (or 1.35 percentage points). This means that, on average, tournament teams attempted 1.35 more free throws per 100 field goal attempts compared to teams that didn’t make the tournament. This statistic helps us understand whether drawing more fouls may correlate with postseason success.

## Permutation Test Function

```{r}
# Function to simulate one permutation of the group labels and return mean difference in FTR
perm_test <- function(data) {
  data |> 
    mutate(shuffled_group = sample(made_tourney)) |> 
    group_by(shuffled_group) |> 
    summarise(mean_ftr = mean(ftr), .groups = "drop") |> 
    pivot_wider(names_from = shuffled_group, values_from = mean_ftr) |> 
    transmute(diff = yes - no) |> 
    pull(diff)
}
```

## Simulate Permutations

```{r}
# Run 1000 permutations and compute one-sided p-value
set.seed(123)
perm_results <- map_dbl(1:1000, ~ perm_test(cbb))

# Testing if tournament teams have higher FTR
p_val <- mean(perm_results >= obs_diff)
p_val
```

## Plot the Null Distribution

```{r}
# Plot the null distribution of differences with the observed difference highlighted
tibble(diff = perm_results) |> 
  ggplot(aes(x = diff)) +
  geom_histogram(binwidth = 0.05, fill = "lightblue", color = "black") +
  geom_vline(xintercept = obs_diff, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test: Free Throw Rate Difference",
    x = "Simulated Difference in Mean FTR",
    y = "Frequency")
```

**Interpretation:**\
This plot shows the distribution of simulated mean differences in Free Throw Rate between tournament and non-tournament teams under the null hypothesis — that there is no real difference. The red dashed line marks the actual observed difference from the real data. Since the red line is far to the right of most of the simulated differences, it suggests that the observed result is unlikely to occur by random chance alone. This visual supports rejecting the null hypothesis.

## Conclusion

Teams that made the NCAA tournament had, on average, a Free Throw Rate about 1.35 percentage points higher than non-tournament teams. Our permutation test produced a p-value below 0.01, indicating this difference is unlikely due to random variation alone. This suggests that getting to the free throw line more often may be a factor in team success.

## Data Source

**Accessed via:**\

Sundberg, A. (2021). *College Basketball Dataset* \[Data set\]. Kaggle. <https://www.kaggle.com/datasets/andrewsundberg/college-basketball-dataset>\

This dataset was compiled and uploaded by Andrew Sundberg to Kaggle to support sports analytics, data exploration, and machine learning applications focused on NCAA Division I men's basketball. The dataset includes team statistics and rankings for multiple seasons.

**Original source:**\

Sports Reference LLC. (n.d.). *College Basketball Statistics*. Sports-Reference.com. <https://www.sports-reference.com/cbb/>\

The original data was collected and published by Sports Reference, an online platform that aggregates historical sports statistics from publicly available NCAA records, box scores, and team reports. The site aims to support research, journalism, and fan-driven sports analysis.
