---
title: "Exploring U.S. Policing Trends Through Data"
author: "Henok Dawite"
date: "April 30, 2025"
format: html
---

**Introduction**

This project explores racial disparities in traffic stop outcomes using data from the Stanford Open Policing Project (Pierson et al., 2020). Using SQL to query a MariaDB database, we investigate differences in search rates across multiple cities and dig deeper into patterns within a single city. Our goal is to surface insights into how policing practices differ by race and geography using fully SQL-based data wrangling and R for visualization.

```{r}
library(DBI)
library(RMariaDB)
library(ggplot2)
library(scales)

# Establish connection to the traffic database
con_traffic <- dbConnect(
  RMariaDB::MariaDB(),
  dbname   = "traffic",
  host     = Sys.getenv("TRAFFIC_HOST"),
  user     = Sys.getenv("TRAFFIC_USER"),
  password = Sys.getenv("TRAFFIC_PWD")
)
```

### Comparing Search Rates Across Cities and Races

```{sql race_search, connection=con_traffic, output.var="race_search"}
SELECT
  city,
  subject_race,
  ROUND(100.0 * AVG(search_conducted), 2) AS search_rate
FROM (
    SELECT 'Charlotte' AS city, subject_race, search_conducted
      FROM nc_charlotte_2020_04_01
      WHERE YEAR(date) BETWEEN 2010 AND 2014
    UNION ALL
    SELECT 'Raleigh' AS city, subject_race, search_conducted
      FROM nc_raleigh_2020_04_01
      WHERE YEAR(date) BETWEEN 2010 AND 2014
    UNION ALL
    SELECT 'Greensboro' AS city, subject_race, search_conducted
      FROM nc_greensboro_2020_04_01
      WHERE YEAR(date) BETWEEN 2010 AND 2014
) AS combined
GROUP BY city, subject_race
ORDER BY city, search_rate DESC;
```

```{r, echo=FALSE}
race_search <- subset(race_search, subject_race != "unknown")
```

```{r}
ggplot(race_search, aes(x = reorder(subject_race, -search_rate),
                        y = search_rate, fill = city)) +
  geom_col(position = "dodge", color = "black") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Search Rates by Race (2010–2014)",
    x = "Race",
    y = "Search Rate (%)",
    fill = "City"
  ) + 
  theme_minimal()
```

**Description:** This is a plot comparing search rates by race for Charlotte, Raleigh, and Greensboro (all North Carolina cities). Charlotte shows that Black drivers have the highest search rate at around 6.75% compared to just over 2.25% of White drivers; Raleigh is second with almost 3.95% for Blacks and 1.95% for Whites; Greensboro, similar to Charlotte, has about 5.14% for Black people and 2.41% for White people. This alludes to a consistent frame across these three cities that no matter where individuals are stopped, historically Black males and females are searched disproportionately higher—and by a long shot—when pulled over in comparison to their White counterparts. The fact that the three cities are so consistent in their gaps makes it hard to disavow random variance or specific variances in one city's policing efforts. However, this represents a national problem that indicates there is a systematic difference between how one race is treated against another. However, it's important to note that these data alone cannot establish causality, as there could be other variables influencing search decisions that aren't captured in this analysis. Hispanic and Asian drivers tend to float around the middle or lower ends of the extremes; for example, Hispanic drivers have a higher overall search percentage than White drivers but lower than Black drivers.

### Stop Outcomes in Oakland, California

```{sql oakland_outcomes, connection=con_traffic, output.var="oakland_outcomes"}
SELECT
  outcome,
  COUNT(*) AS count,
  ROUND(100.0 * COUNT(*) / total.total_count, 2) AS percent
FROM ca_oakland_2020_04_01,
     (SELECT COUNT(*) 
     AS total_count FROM ca_oakland_2020_04_01 WHERE outcome IS NOT NULL)
     AS total
WHERE outcome IS NOT NULL
GROUP BY outcome, total.total_count
ORDER BY count DESC;
```

```{r}
ggplot(oakland_outcomes, aes(x = reorder(outcome, -count), y = percent)) +
  geom_col(fill = "lightblue", color = "black") +
  labs(
    title = "Stop Outcomes in Oakland (2013-2017)",
    x = "Stop Outcome",
    y = "Percentage of Total Stops"
  ) + 
  theme_minimal()
```

**Description:** This is the outcome of traffic stops in Oakland, California. For the total number of traffic stops, 52% end in citation, 31% end in warning, and 17% end in arrest. The fact that over half of all traffic stops end in citation shows that the majority of traffic violations are minor in nature, yet Oakland law enforcement officials felt that these actions were egregious enough to render a formal citation. With almost one-third of the actions resulting in a warning, this could either mean that the officers give a lot of warnings based on their discretion or that the action was so minor that a citation would be an overreach; either way, this is a fair amount of leeway given. The most surprising statistic is that 17% end in arrest which means that 1 out of every 6 people gets arrested during a stop. This is a relatively high number for either having a large amount of stops based on reasonable suspicion or, conversely, having an overly aggressive policy towards law enforcement where more arrests are had than in other precincts.

**Conclusion:** Using SQL and R, we analyzed search rates and stop outcomes across and within cities. We used multiple SQL tables and functions (`UNION ALL`, `ROUND`, `CASE WHEN`, `GROUP BY`, `ORDER BY`, `WHERE`) to handle our wrangling exclusively within the database. This approach allowed us to uncover patterns of racial disparity and enforcement trends across jurisdictions. Our comparison across cities revealed that Black drivers face disproportionately high search rates, while city-level breakdowns like in Oakland showed that most stops result in either citations or warnings, but that arrests are not uncommon. The data were sourced from the Stanford Open Policing Project (Pierson et al., 2020), and the analysis suggests that both the rate and nature of stops vary substantially depending on location and race.

**Citations**

**Academic Source:**\

Pierson, E., Corbett-Davies, S., & Goel, S. (2020). *A large-scale analysis of racial disparities in police stops across the United States*. *Nature Human Behaviour*, 4(7), 736–745. <https://doi.org/10.1038/s41562-020-0858-1>

**Dataset Source:**\

Stanford Open Policing Project. (n.d.). *Traffic stop data from across the United States*. Stanford Computational Policy Lab. Retrieved from <https://openpolicing.stanford.edu/data/>

The Stanford Open Policing Project collects, cleans, and standardizes traffic stop data from law enforcement agencies across the U.S. to enable analyses of racial disparities in policing practices. The data includes over 200 million records from state patrol and municipal police departments.
