---
title: "Analysis of Netflix Movies and TV Shows"
description: |
  This is an analysis of Netflix's content catalog using R's 'tidyverse' package. 
  Visualizations include a bar chart showing the number of titles added per year, 
  a box plot comparing description word counts by content type, and an analysis 
  identifying the most frequent words following 'The' in titles, presented in a 
  horizontal bar chart. This analysis offers insights into content trends, 
  description characteristics, and title naming conventions on Netflix.
author: Henok Dawite
date: March 4, 2025
format: html
execute:
  warning: false
  message: false
toc: false
about:
  template: trestles
  links: 
    - text: "Netflix Dataset"
      href: "https://www.kaggle.com/datasets/shivamb/netflix-shows"
---

```{r}
library(tidyverse)

# Load the dataset
netflix_data <- read.csv("netflix_titles 2.csv", stringsAsFactors = FALSE)

# Data Cleaning
netflix_data <- netflix_data |>
  mutate(
    date_added = as.Date(date_added, format="%B %d, %Y"),
    title = as.character(title)
  )
```

```{r}
# Extract year from date_added for visualization
netflix_data_clean <- netflix_data |>
  mutate(year_added = str_extract(as.character(date_added), "\\d{4}")) |>
  filter(!is.na(year_added))

# Number of titles per year
p1 <- ggplot(netflix_data_clean, aes(x = year_added)) +
  geom_bar(fill = "steelblue", color = "black", width = 0.8) +
  labs(title = "Number of Titles Added Per Year", x = "Year", y = "Count")

print(p1)
```

This is a bar chart that indicates how many titles were added to the Netflix library each year. If we determine that more titles were added one year than the next, we can deduce whether Netflix was in a position to try to reduce titles at one point due to business constraints or, conversely, in a stable position to assess market forces or consumer demand and expand availability.

```{r}
# Count the number of words in each description for visualization
netflix_data <- netflix_data |>
  mutate(word_count = str_count(description, "\\w+"))

# Average word count in description by type
p2 <- ggplot(netflix_data, aes(x = type, y = word_count, fill = type)) +
  geom_boxplot() +
  labs(title = "Word Count in Description by Type", x = "Type", y = "Word Count") +
  scale_fill_brewer(palette = "Set3")

print(p2)
```

This box plot displays the distribution of word counts in content descriptions, categorized by movies and TV shows. Looking at the actual data, we can see that the distributions are quite similar between the two content types. Both movies and TV shows have median description lengths of approximately 23-25 words, with similar interquartile ranges but Tmay have a slightly wider spread, indicating more variability in description lengths. This suggests that Netflix maintains relatively consistent description lengths across content types, likely adhering to internal style guidelines for content descriptions regardless of whether the title is a movie or TV show.

```{r}
# Extract word after "The" in titles
netflix_data <- netflix_data |>
  mutate(word_after_the = str_extract(title, "(?<=The )\\w+"))

# Count top 10 most common words
word_after_the_counts <- netflix_data |>
  drop_na(word_after_the) |>
  count(word_after_the, name = "count") |>
  arrange(desc(count)) |>
  slice_head(n = 10)

# Horizontal bar chart
p3 <- word_after_the_counts |>
  ggplot(aes(x = reorder(word_after_the, count), y = count, fill = count)) +
  geom_bar(stat = "identity", color = "black") +
  coord_flip() +
  labs(title = "Top 10 Words Following 'The' in Netflix Titles", x = "Word", y = "Count") +
  scale_fill_gradient(low = "moccasin", high = "darkorange") +
  theme(legend.position = "none")

print(p3)
```

This horizontal bar chart showcases the ten most common words that immediately follow 'The' in Netflix titles. This chart analyzes the frequency of words that commonly follow the word "The" in titles on Netflix. The most frequent word is "Last," followed by "Movie," "Series," "Great," "Story," "Legend," "Beginning," "New," "Little," and "Devil." This data suggests that Netflix titles often use these words to create intrigue or convey the theme of the content. For instance, words like "Last," "Great," and "Legend" might indicate titles with epic or significant narratives, while "Movie" and "Series" are straightforward descriptors of the format.

Data Source:

**Accessed via:**\

Shivam Bansal. (2019). *Netflix Movies and TV Shows Dataset*. Kaggle. Retrieved from: <https://www.kaggle.com/datasets/shivamb/netflix-shows>

**Original Data Source:**\

Collected from Netflix’s publicly available catalog data. This dataset includes TV shows and movies available on Netflix as of 2021, with metadata such as title, director, cast, release year, country, date added, rating, and description.
